PyPy's Approach to Virtual Machine Construction

### 2006-07-12 talk/dls2006/paper.tex
pointer to code repository (we are again above 12 pages, something will have to give)
clarify interpreter line count and role.

There are two major components in PyPy:
%
\begin{enumerate}
\item the \textit{Standard Interpreter}: an implementation of the Python programming
language, mostly complete and compliant with the current version of the
language, Python 2.4.
\item the \textit{Translation Process}: a translation tool-suite whose goal is to
compile subsets of Python to various environments.
\end{enumerate}
%
In particular, we have defined a subset of the Python language called
``restricted Python'' or RPython.  This sublanguage is not restricted
syntactically, but only in the way it manipulates objects of different
types.  The restrictions are a compromise between the expressivity and
the need to statically infer enough type information to generate efficient code.
The foremost purpose of the translation tool-suite is to compile such
RPython programs to a variety of different platforms.

Our current efforts, and the present paper, focus on this tool-suite.
We will not describe the Standard Interpreter component of PyPy in the
sequel, other than to mention that it is written in RPython and can thus be
translated.  At close to 90,000 lines of code, it is the largest RPython
program that we have translated so far.  More information can be found
in \cite{A}.

@@

The numbers in parenthesis are slow-down factors compared to CPython.
These measures reflect PyPy revision 27815, compiled with GCC 3.4.4.
LLVM is version 1.8cvs (May 11, 2006).  The machine runs GNU/Linux SMP
on an Intel(R) Pentium(R) 4 CPU at 3.20GHz with 2GB of RAM and 1MB of
cache.  The rows correspond to variants of the translation process, as
follows:

>>>>

There are two major components in PyPy:
%
\begin{enumerate}
\item the \textit{Standard Interpreter}: an implementation of the Python programming
language, mostly complete and compliant with the current version of the
language, Python 2.4.
\item the \textit{Translation Process}: a translation tool-suite whose goal is to
compile subsets of Python to various environments.
\end{enumerate}
%
In particular, we have defined a subset of the Python language called
``restricted Python'' or RPython.  This sublanguage is not restricted
syntactically, but only in the way it manipulates objects of different
types.  The restrictions are a compromise between the expressivity and
the need to statically infer enough type information to generate efficient code.
The purpose of the translation tool-suite is to compile such
RPython programs to a variety of different platforms.

Our current efforts, and the present paper, focus on this tool-suite.
We will not describe the Standard Interpreter component of PyPy in the
sequel, other than to mention that it is written in RPython and can thus be
translated.  At close to 90,000 lines of code\footnote{The core interpreter
is more like 30,000 lines of code, the rest includes built-in modules and 30,000 lines
of auto-generated unicode character database data.}, it is the largest RPython
program that we have translated so far and is the main target driving
the development of the tool-chain.  More information can be found
in \cite{A}.

@@

The numbers in parenthesis are slow-down factors compared to CPython.
These measures reflect PyPy revision 27815\footnote{PyPy is an open-source project
under the MIT license, PyPy source repository using subversion lives at 
http://codespeak.net/svn/pypy/dist/.}, 
compiled with GCC 3.4.4.  LLVM is version 1.8cvs (May 11, 2006).  The
machine runs GNU/Linux SMP on an Intel(R) Pentium(R) 4 CPU at 3.20GHz
with 2GB of RAM and 1MB of cache.  The rows correspond to variants of
the translation process, as follows:
